The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-30 manuscript/5pillars.Rmd
improvements to fig 1, fig 4

Computational reproducibility is a simple premise in theory, but is difficult to
achieve in practice.
Building upon past efforts and proposals to maximize reproducibility and rigor in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) code version control and sharing, (2) compute environment control,
(3) literate programming, (4) documentation and (5) FAIR data sharing.
These best practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatic data analysts, but should be relevant for other
domains.

@@

The ramifications of irreproducible and unreliable research includes misleading the community,
wasting research funds, slowing scientific progress, eroding public confidence in science and
tarnishing the reputation of associated institutions and colleagues.
In certain cases, irreproducible bioinformatics has the potential to place patient safety at risk.

@@

text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
text( 5 , 0.2 , lab="End-to-end automated process",cex=1.2)
text(5,7,lab="Process validation & testing",cex=1.2)

@@

polygon(c(6,11.5,11.5,6),c(0,0,0.9,0.9), col=cols2[1], border = NA)
polygon(c(6,11.5,11.5,6),c(1,1,1.9,1.9), col=cols2[3], border = NA)

labs3 <- c("Hardware","Hypervisor")
text(c(8.5,8.5),c(0.4,1.4),labels=labs3)

polygon(c(6,7.8,7.8,6),c(2,2,4.5,4.5), col=cols2[2], border = NA)
polygon(c(8,9.6,9.6,8),c(2,2,4.5,4.5), col=cols2[2], border = NA)
polygon(c(9.8,11.5,11.5,9.8),c(2,2,4.5,4.5), col=cols2[2], border = NA)

text(c(6.85,8.8,10.65),c(3.5,3.5,3.5),labels="Guest\nOperating\nSystem")

@@

### Process validation and testing

@@

Literate programming also provides an opportunity to record sanity checks in an output report.
In R, commands like `dim()`, `length()`, `head()`, `str()` are a simple way to show the features of
a dataset to validate that data transformation steps are working as expected.

@@

In software development, continuous integration/continuous delivery (CI/CD) is a process of automated
software compilation and testing.
A derivation of this for data analysis has been termed "continuous analysis", and involves automatic
execution and testing whenever a change is made to a project's code base
[@Beaulieu-Jones2017-vj; @Blischak2019-dq].
If the workflow executes successfully and tests are passed, the repository and linked research article
can be decorated with a "badge" to inform users of the lack of errors in the code,
similar to badges used for other Open Science practices [@Kidwell2016-bb].
If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
article can be assessed computationally, then such tests could be used to demonstrate computational
reproducibility without the need for individuals to execute the workflow themselves [@Suetake2022-zq].

@@

Another hindrance to reproducibility more broadly is the fixation on novel research directions rather
than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
@Munafo2017-ty; @Diong2021-is].
This is a direct result of research assessment policies that incentivize journal-level prestige
metrics and number of publications rather than rigor [@Munafo2017-gf; @Oransky2023-zb].
A change in research assessment ("eg: track-record") that recognizes and credits contributions toward
reproducibility and rigor would help decrease funds wasted on sloppy research.

>>>>

Computational reproducibility is a simple premise in theory, but is difficult to
achieve in practice.
Building upon past efforts and proposals to maximize reproducibility and rigor in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) literate programming, (2) code version control and sharing,
(3) compute environment control, (4) FAIR and persistent data sharing and (5) documentation.
These practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatic data analysts, but should be relevant for other
domains.

@@

The ramifications of irreproducible and unreliable research includes misleading the community,
wasting research funds, slowing scientific progress, eroding public confidence in science and
tarnishing the reputation of associated institutions and colleagues.
In clinical research, irreproducible bioinformatics has the potential to place patient
safety at risk.

@@

text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
text( 5 , 0.2 , lab="End-to-end automated process",cex=1.2)
text(5,7,lab="Continuous validation",cex=1.2)

@@

polygon(c(6,11.5,11.5,6),c(0,0,0.9,0.9), col=cols2[1], border = NA)
polygon(c(6,11.5,11.5,6),c(1,1,1.9,1.9), col=cols2[2], border = NA)
polygon(c(6,11.5,11.5,6),c(2,2,2.9,2.9), col=cols2[3], border = NA)

labs3 <- c("Hardware","Host Operating System","Hypervisor")
text(c(8.5,8.5,8.5),c(0.4,1.4,2.4),labels=labs3)

polygon(c(6,7.8,7.8,6),c(3,3,4.5,4.5), col=cols2[2], border = NA)
polygon(c(8,9.6,9.6,8),c(3,3,4.5,4.5), col=cols2[2], border = NA)
polygon(c(9.8,11.5,11.5,9.8),c(3,3,4.5,4.5), col=cols2[2], border = NA)

text(c(6.85,8.8,10.65),c(3.8,3.8,3.8),labels="Guest\nOperating\nSystem")

@@

### Continuous validation

@@

Literate programming also provides an opportunity to record sanity checks in an output report.
In R, commands like `dim()`, `length()`, `head()`, `str()` are a simple way to show the features of
a dataset to validate that data transformation steps are working as expected.
Histograms, boxplots, scatterplots and other simple charts are similarly effective.

@@

In software development, continuous integration/continuous delivery (CI/CD) is a process of automated
software compilation and testing.
A derivation of this for data analysis has been termed "continuous analysis", and involves automatic
execution and testing whenever a change is made to a project's code base
[@Beaulieu-Jones2017-vj; @Blischak2019-dq].
We prefer the term "continuous validation" as it places more emphasis on the correctness of analyses
to ensure they are fit to share.
If a workflow is executed successfully and tests are passed, the repository and linked research article
can be decorated with a "badge" to inform users of the lack of errors in the code,
similar to badges used for other Open Science practices [@Kidwell2016-bb].
If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
article can be assessed computationally, then such tests could be used to demonstrate computational
reproducibility without the need for individuals to execute the workflow themselves [@Suetake2022-zq].

@@

Another hindrance to reproducibility more broadly is the fixation on novel research directions rather
than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
@Munafo2017-ty; @Diong2021-is].
This is a direct result of research assessment policies that incentivize journal-level prestige
metrics and number of publications rather than rigor [@Munafo2017-gf; @Oransky2023-zb].
A change in research assessment ("eg: track-record") that recognizes and credits contributions toward
reproducibility and rigor would help decrease funds wasted on sloppy research.
Such changes would be welcomed by industry as they would face fewer replication failures
[@Rosenblatt2016-cq], which could speed development of research into useful products.
